---
title: "Exploratory Data Analysis - Week 1 assignment"
output: html_notebook
---

This notebook has the work for the Exploratory Data Analysis week 1. To conform to the submission requirements, it needs to be converted into individual R scripts for each image that is produced.


Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.


Download the source data:
```{r download}
localfile <- "power_consumption.zip"
url <- "https://d396qusza40orc.cloudfront.net/exdata%2Fdata%2Fhousehold_power_consumption.zip"
dir <- "data"
extracted_filename = "household_power_consumption.txt"
# Download and unzip only if the file isn't already local. 
if(!file.exists(localfile)){
  download.file(url,localfile, mode = "wb") # use binary mode since it's a zip file
}
if(!file.exists(dir)){
  unzip(localfile, files = NULL, exdir=".")
}
```
Load the extracted data file and condition it. We'll add a timestamp column for convenience of generating the plots. 
```{r load data}
pwr <- read.table(extracted_filename, sep=";", header=TRUE, colClasses="character")
pwr$timestamp = paste(pwr$Date, pwr$Time, sep=' ')
pwr$timestamp = strptime(pwr$timestamp,format ='%d/%m/%Y %H:%M:%S')                      
```

The first plot is a frequency plot of Global active power vs frequency for a specified date (2007-02-02) 
```{r plot 1}
subpwr <- subset(pwr, Date=="2/2/2007" | Date=="1/2/2007") # we could also convert to dates and do arithmetic, but this string comparison is much faster.
h <- hist(
  as.numeric(subpwr$Global_active_power), 
  col="red", 
  main="Global Active Power",
  xlab="Global active power (kilowatts)"
     )
dev.copy(png, file="plot1.png")
dev.off()
```


The second plot is a frequency plot of GAP (y) vs timestamped data(x)

```{r plot 2}
plot(subpwr$timestamp, subpwr$Global_active_power, 
     type = 'l',
     main = 'Global active power', 
     xlab = NA,
     ylab = 'Global active power (kilowatts)'
)
dev.copy(png, file="plot2.png")
dev.off()
```
The third plot shows three sub metering values on our two-day period of data. 
```{r plot 3}

plot(subpwr$timestamp, subpwr$Sub_metering_1, type="l",ylab = "Energy sub metering", xlab="" , col="black")
points(subpwr$timestamp, subpwr$Sub_metering_2, type="l", col="red")
points(subpwr$timestamp, subpwr$Sub_metering_3, type="l", col="blue")

# add a legend
legend("topright", 
       lty = c(1, 1, 1), # line type 1 is a solid line. Required argumen 
       col = c("black", "red", "blue"),  # colors
       legend = c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3")
       )

# output PNG file
dev.copy(png, file="plot3.png")
dev.off()
```

Finally we combine all the plots above
```{r plot 4}


par(mfrow=c(2,2))

with(subpwr, {
  plot(timestamp, Global_active_power, type="l", ylab="Global Active Power", xlab="")
  plot(timestamp, Voltage, type="l", ylab="Voltage", xlab="datetime")
  plot(timestamp, Sub_metering_1, type="l",ylab = "Energy sub metering", xlab="" , col="black")
points(timestamp, Sub_metering_2, type="l", col="red")
points(timestamp, Sub_metering_3, type="l", col="blue")
legend("topright", 
       lty = c(1, 1, 1), # line type 1 is a solid line. Required argument 
       col = c("black", "red", "blue"),  # colors
       legend = c("Sub_metering_1", "Sub_metering_2", "Sub_metering_3"),
       bty="n",
       cex=.7, xjust=1, yjust=1
       )
plot(timestamp, Global_reactive_power, type="l", ylab="Global_reactive_power", xlab="datetime")
})

# output PNG file
dev.copy(png, file="plot3.png",width = 480, height = 480)
dev.off()
```

